---
title: "HutanDesa_Vegetation"
author: "Gene Estrada"
date: "2023-10-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
setwd("~/HutanDesa_Biodiversity/vegetation")

library(tidyverse)
library(vegan)

#reading in vegetation data from 2016 HD survey 
v <- read.csv(file = "Veg_2016.csv", header = TRUE, na.strings = c("", " "))

#reading in vegetation data from 2021 
v21 <- read.csv(file = "Veg_2021.csv", header = TRUE, na.strings = c("", " "))
loc <- read.csv(file = "PlotsCoding_2021.csv", header = TRUE)

v21 <- left_join(v21, loc[,c("survey.area", "transect", "plot", "loc")], by = "plot")

v21$loc[v21$loc == "nipah.kuning"] <- "Nipah Kuning"
v21$loc[v21$loc == "padu.banjar"] <- "Padu Banjar"
v21$loc[v21$loc == "pemangkat"] <- "Pemangkat"
v21$loc[v21$loc == "penjalaan"] <- "Penjalaan"
v21$loc[v21$loc == "pulau.kumbang"] <- "Pulau Kumbang"
v21$loc[v21$loc == "rantau.panjang"] <- "Rantau Panjang"

#reading in Andy's pheno data 
pheno <- read.csv(file = "Jul2015-Sep2020_pheno_AJM.csv", header = TRUE)
pheno$hab_unbiased <- factor(pheno$hab_unbiased, levels = c("PS", "FS", "AB", "LS", "LG", "UG"))

#replacing "NULL" and other odd values in the Species and Genus column with NA
#table not in the best shape, many species spelled slightly different, will inflate species counts
pheno$species[pheno$species == "NULL"] <- NA
pheno$species[pheno$species == ""] <- NA
pheno$species[pheno$species == "`"] <- NA
pheno$genus[pheno$genus == ""] <- NA
pheno$genus[pheno$genus == "?"] <- NA
pheno$genus[pheno$genus == "???"] <- NA
pheno$genus[pheno$genus == "????"] <- NA
pheno$genus[pheno$genus == "NA"] <- NA

#fixing some of the names manually - some I'm not sure what to do with
pheno$species <- gsub("\\?", "", as.character(pheno$species))
pheno$species[pheno$species == "zwageri"] <- "zwagerii"
pheno$species[pheno$species == "zwageriii"] <- "zwagerii"
pheno$species[pheno$species == "stelllatus"] <- "stellatus"
pheno$species[pheno$species == "sp nov"] <- "spnov"
pheno$genus[pheno$genus == "uvaria"] <- "Uvaria"

#creating a column for Latin Binomial for rows that have both Genus and Species ID's
pheno$scientific.name <- ifelse(!is.na(pheno$genus) & !is.na(pheno$species), 
                            paste(pheno$genus, pheno$species, sep = " "), 
                            NA)

#creating a column for Latin Binomial for rows that have both Genus and Species ID's
v$scientific.name <- ifelse(!is.na(v$Genus) & !is.na(v$Species), 
                            paste(v$Genus, v$Species, sep = " "), 
                            NA)

v21$scientific.name <- ifelse(!is.na(v21$genus) & !is.na(v21$species), 
                            paste(v21$genus, v21$species, sep = " "), 
                            NA)
```

### Exploration

Starting with the 2016 vegetation data, which covers four Hutan Desas: Nipah Kuning, Padu Banjar, Pamangkat, and Pulau Kumbang.
```{r}
#it looks like all stems except 34 were ID'd to the species level
table(is.na(v$Species))

#only one stem did not have an ID to the genus level
table(is.na(v$Genus))
```


The 2021 surveys cover the same HD sites as above, plus Penjalaan and Rantau Panjang
```{r}
#all but 46 stems were ID'd to the species level
table(is.na(v21$species))

#all were ID'd to the genus level
table(is.na(v21$genus))
```


### Vegetation Diversity

#### 2016 data - 
Table showing counts of plant species by Hutan Desa 
```{r}
species_table <- v %>%
  group_by(Location, scientific.name) %>%
  filter(!is.na(scientific.name)) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = Location, values_from = count, values_fill = 0) %>%
  arrange(desc(`Nipah Kuning`))

print(species_table, n = nrow(species_table))  
```


Table showing counts of plant genera by Hutan Desa
```{r}
genus_table <- v %>%
  group_by(Location, Genus) %>%
  filter(!is.na(Genus)) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = Location, values_from = count, values_fill = 0) %>%
  arrange(desc(`Nipah Kuning`))

print(genus_table, n = nrow(genus_table))  
```


Comparing plant species composition and diversity by Hutan Desa
```{r}
# Calculate species richness by site
species_richness <- v %>%
  group_by(Location) %>%
  summarise(SpeciesRichness = length(unique(scientific.name))) %>%
  arrange(desc(SpeciesRichness))

# Calculate Shannon diversity index by site
shannon_index <- diversity(table(v$Location, v$scientific.name), index = "shannon")
shannon_index <- rownames_to_column(as.data.frame(shannon_index), var = "Location")

# Calculate Simpson diversity index by site
simpson_index <- diversity(table(v$Location, v$scientific.name), index = "simpson")
simpson_index <- rownames_to_column(as.data.frame(simpson_index), var = "Location")

#create summary table
veg_summary <- merge(species_richness, shannon_index, by = "Location")
veg_summary <- merge(veg_summary, simpson_index, by = "Location")
veg_summary
```

Comparing plant genus composition and diversity by Hutan Desa
```{r}
#by genus
genus_richness <- v %>%
  group_by(Location) %>%
  summarise(GenusRichness = n_distinct(Genus)) %>%
  arrange(desc(GenusRichness))

#Calculate Shannon diversity index by site
shannon_index_gen <- diversity(table(v$Location, v$Genus), index = "shannon")
shannon_index_gen <- rownames_to_column(as.data.frame(shannon_index_gen), var = "Location")

#Calculate Simpson diversity index by site
simpson_index_gen <- diversity(table(v$Location, v$Genus), index = "simpson")
simpson_index_gen <- rownames_to_column(as.data.frame(simpson_index_gen), var = "Location")

#create summary table
veg_summary_gen <- merge(genus_richness, shannon_index_gen, by = "Location")
veg_summary_gen <- merge(veg_summary_gen, simpson_index_gen, by = "Location")
veg_summary_gen
```


#### 2021 data - 
Table showing counts of plant species by Hutan Desa 
```{r}
species_table_21 <- v21 %>%
  group_by(loc, scientific.name) %>%
  filter(!is.na(scientific.name)) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = loc, values_from = count, values_fill = 0) %>%
  arrange(desc(`Nipah Kuning`))

print(species_table_21, n = nrow(species_table_21))  
```

Table showing counts of plant genera by Hutan Desa
```{r}
genus_table_21 <- v21 %>%
  group_by(loc, genus) %>%
  filter(!is.na(genus)) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = loc, values_from = count, values_fill = 0) %>%
  arrange(desc(`Nipah Kuning`))

print(genus_table_21, n = nrow(genus_table_21))  
```

Comparing plant species composition and diversity by Hutan Desa
```{r}
# Calculate species richness by site
species_richness <- v21 %>%
  group_by(loc) %>%
  summarise(SpeciesRichness = length(unique(scientific.name))) %>%
  arrange(desc(SpeciesRichness))

# Calculate Shannon diversity index by site
shannon_index <- diversity(table(v21$loc, v21$scientific.name), index = "shannon")
shannon_index <- rownames_to_column(as.data.frame(shannon_index), var = "loc")

# Calculate Simpson diversity index by site
simpson_index <- diversity(table(v21$loc, v21$scientific.name), index = "simpson")
simpson_index <- rownames_to_column(as.data.frame(simpson_index), var = "loc")

#create summary table
veg_summary_21 <- merge(species_richness, shannon_index, by = "loc")
veg_summary_21 <- merge(veg_summary_21, simpson_index, by = "loc")
veg_summary_21
```


Ordination analysis - ***still working on this
```{r}
#get data in right format - I think I need to not group by site here, and by survey/transect instead
t1 <- v %>%
  group_by(Location, scientific.name) %>%
  filter(!is.na(scientific.name)) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = scientific.name, values_from = count, values_fill = 0)

#run ordination
MDS1 <- metaMDS(t1[,-1], distance = "euclidean")

#plot
plot(MDS1, type = "points")
ordihull(MDS1, groups = 1:4, col = "green", draw = "polygon")
```


#### Andy's pheno data from CPRS

Creating list of species in pheno plots with counts. Note, pheno plots were not placed in the Montane (MO) forest, so there is no data here to compare for the montane. 

Not many species are listed here. Looking at the data, it looks like not many stems were identified to the species level.
```{r}
species_table_CPRS <- pheno %>%
  filter(hab_unbiased == "PS" & pheno_year == "2016") %>%
  group_by(scientific.name) %>%
  filter(!is.na(scientific.name)) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(species_table_CPRS, n = nrow(species_table_CPRS))  
```

Doing the same thing again but for genus. Many more stems were identified to the genus level. 
```{r}
genus_table_CPRS <- pheno %>%
  filter(hab_unbiased == "PS" & pheno_year == "2016") %>%
  group_by(genus) %>%
  filter(!is.na(genus)) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(genus_table_CPRS, n = nrow(genus_table_CPRS))  
```

Calculating plant species diversity indices by forest type at CPRS
```{r}
# Calculate species richness by site
species_richness_CPRS <- pheno %>%
  group_by(hab_unbiased) %>%
  summarise(SpeciesRichness = n_distinct(scientific.name)) %>%
  arrange(desc(SpeciesRichness))

# Calculate Shannon diversity index by site
shannon_index_CPRS <- diversity(table(pheno$hab_unbiased, pheno$scientific.name), index = "shannon")
shannon_index_CPRS <- rownames_to_column(as.data.frame(shannon_index_CPRS), var = "hab_unbiased")

# Calculate Simpson diversity index by site
simpson_index_CPRS <- diversity(table(pheno$hab_unbiased, pheno$scientific.name), index = "simpson")
simpson_index_CPRS <- rownames_to_column(as.data.frame(simpson_index_CPRS), var = "hab_unbiased")

#create summary table
veg_summary_CPRS <- merge(species_richness_CPRS, shannon_index_CPRS, by = "hab_unbiased")
veg_summary_CPRS <- merge(veg_summary_CPRS, simpson_index_CPRS, by = "hab_unbiased")

veg_summary_CPRS %>%
  arrange(desc(hab_unbiased))
```

Calculating plant genus diversity indices by forest type at CPRS. It might make sense to use these indices for comparison to the Hutan Desa, at least from Andy's data, given the lack of species ID's, as seen above. 
```{r}
# Calculate species richness by site
genus_richness_CPRS <- pheno %>%
  group_by(hab_unbiased) %>%
  summarise(GenusRichness = n_distinct(genus))

# Calculate Shannon diversity index by site
shannon_index_CPRS_gen <- diversity(table(pheno$hab_unbiased, pheno$genus), index = "shannon")
shannon_index_CPRS_gen <- rownames_to_column(as.data.frame(shannon_index_CPRS_gen), var = "hab_unbiased")

# Calculate Simpson diversity index by site
simpson_index_CPRS_gen <- diversity(table(pheno$hab_unbiased, pheno$genus), index = "simpson")
simpson_index_CPRS_gen <- rownames_to_column(as.data.frame(simpson_index_CPRS_gen), var = "hab_unbiased")

#create summary table
veg_summary_CPRS_gen <- merge(genus_richness_CPRS, shannon_index_CPRS_gen, by = "hab_unbiased")
veg_summary_CPRS_gen <- merge(veg_summary_CPRS_gen, simpson_index_CPRS_gen, by = "hab_unbiased")

veg_summary_CPRS_gen %>%
  arrange(desc(hab_unbiased))
```


Comparing plant species diversity between CPRS forests and HD sites. Again, if using Andy's data moving forward in any way, comparing these values at the genus level might be best. 
```{r}
#by species
colnames(veg_summary_CPRS) <- c("Location", "SpeciesRichness", "shannon_index", "simpson_index")
veg_summary_CPRS <- veg_summary_CPRS %>% arrange(Location)
rbind(veg_summary, veg_summary_CPRS)

#by genus
colnames(veg_summary_CPRS_gen) <- c("Location", "GenusRichness", "shannon_index_gen", "simpson_index_gen")
veg_summary_CPRS_gen <- veg_summary_CPRS_gen %>% arrange(Location)
rbind(veg_summary_gen, veg_summary_CPRS_gen)
```


### Structural Comparison

Using structural metrics DBH and Tree height to compare Hutan Desa sites. There doesn't seem to be much difference here. I'm not sure that I'll be using these metrics moving forward, so just visually assessing for now. 

#### 2016 data - 
Tree DBH - one obvious outlier here in the Pemangkat site
```{r}
ggplot(v, aes(x = DBH, y = Location)) +
  geom_violin() + 
  geom_boxplot(width=0.3, alpha=0.2) +
  geom_jitter(height = .1, alpha = 0.2) + 
  labs(title = "Tree Diameters at Hutan Desa Sites (2016)",
       x = "DBH (cm)", y = "") +
  theme_minimal()
```

Tree height - no obvious differences here
```{r}
ggplot(v, aes(x = Tree.Height, y = Location)) +
  geom_violin() + 
  geom_boxplot(width=0.3, alpha=0.2) +
  geom_jitter(height = .1, alpha = 0.2) + 
  labs(title = "Tree Height at Hutan Desa Sites (2016)",
       x = "height (m)", y = "") +
  theme_minimal()
```

#### 2021 data - only has DBH
```{r}
ggplot(v21, aes(x = dbh, y = loc)) +
  geom_violin() + 
  geom_boxplot(width=0.3, alpha=0.2) +
  geom_jitter(height = .1, alpha = 0.2) + 
  labs(title = "Tree Diameters at Hutan Desa Sites (2021)",
       x = "DBH (cm)", y = "") +
  theme_minimal()
```
